---
title: "GBIF test script"
author: "Brady Gates"
date: "09/18/2023"
output: word_document
---

sets up R workspace and loads packages

```{r}

#clears workspace
rm(list=ls())

#sets working directory
setwd("C:/maxent")

#ensures java (jvm.dll) can be found
Sys.setenv(JAVA_HOME='/Library/Internet Plug-Ins/JavaAppletPlugin.plugin/Contents/Home/')
require(rJava)

#gives more memory to java to run bigger models
options(java.parameters = "-Xmx5g")

#loads in necessary packages
library("dismo")
library("usdm")
library("dplyr")

```
```{r}
install.packages("dismo")
install.packages("usdm")
install.packages("dpylr")

```


```{r}
install.packages("raster")
library("raster")
```

function to stack all rasters in an indicated directory

```{r}

stack_folder = function(filepath) {
  
  #gets filepaths of all the tif files in the given folder
  tifdata = list.files(path=filepath, pattern="*.tif")
  filepaths = paste(filepath, tifdata, sep="")

  #stacks all tif files
  wcrasters = stack(filepaths)
  
}

```

function to combine two "wrap-around" rasters
note that the final raster will have extent_a on the left and extent_b on the right
also, extent_a and extent_b should have the same y dimensions
in our case, creates a new raster from worldclim data that continually goes from Asia to North America

```{r}

raster_wrap_comb = function(extent_a, extent_b, raster) {
  
  #creates new extents based off a and b
  ea_shift = extent(0, extent_a[2]-extent_a[1], extent_a[3], extent_a[4])
  eb_shift = extent(ea_shift[2], ea_shift[2]+(extent_b[2]-extent_b[1]), extent_b[3], extent_b[4])
  
  #crops the raster to the extent and sets new extents
  ea_crop = setExtent(crop(raster, extent_a), ea_shift, keepres=FALSE, snap=FALSE)
  eb_crop = setExtent(crop(raster, extent_b), eb_shift, keepres=FALSE, snap=FALSE)

  #return merged rasters
  merge(ea_crop, eb_crop)
  
}

```

loads in presence data

```{r}

#loads in presence data (collected from gbif) 
setwd("C:/maxent/")
occurence = read.csv("C:/maxent/DIVI Cleaned.csv")

#use head to ensure data looks ok
head(occurence)

#trims to just latitude and longitude and
#use head to ensure data is just xy-coordinates 
#should be two coordinates listed as (lon, lat)
occurence = occurence[c("decimalLongitude", "decimalLatitude")]
head(occurence)

```

loads in climate data

```{r}

#gets present climate data directly from worldclim
wccurr = getData('worldclim', var='bio', res=2.5)
#cclgmbi_2-5m/

#gets lgm and future projection rasters from file
wclgm = stack_folder("C:/maxent/sangmodels/cclgmbi_2-5m/")
wcrcp85 = stack_folder("C:/maxent/sangmodels/cc85bi70/")
wcrcp60 = stack_folder("C:/maxent/sangmodels/cc60bi70/")
wcrcp45 = stack_folder("C:/maxent/sangmodels/cc45bi70/")
wcrcp26 = stack_folder("C:/maxent/sangmodels/cc26bi70/")

#extract current conditions at occurence points
wccurr_vals = extract(wccurr, occurence, cellnumbers=TRUE)

#removes duplicated rows from occurence data
occ_distinct = subset(distinct(na.omit(cbind(as.data.frame(wccurr_vals), occurence)), cells, .keep_all= TRUE), select=-c(1:20))

#removes duplicated rows from extracted values
wccurr_distinct = subset(distinct(na.omit(as.data.frame(wccurr_vals)), cells, .keep_all= TRUE), select=-c(cells))

#uses VIF to exclude correllated variables with a threshold of 10
vcurr = exclude(wccurr_distinct, vifstep(wccurr_distinct, th=10))

#gets only uncorrellated rasters
wccurr_uncorr = subset(wccurr, colnames(vcurr))
wclgm_uncorr = subset(wclgm, colnames(vcurr))
wcrcp85_uncorr = subset(wcrcp85, colnames(vcurr))
wcrcp60_uncorr = subset(wcrcp60, colnames(vcurr))
wcrcp45_uncorr = subset(wcrcp45, colnames(vcurr))
wcrcp26_uncorr = subset(wcrcp26, colnames(vcurr))

#ensures all names match up
print(rbind(colnames(vcurr), names(wccurr_uncorr), names(wclgm_uncorr), names(wcrcp85_uncorr), names(wcrcp60_uncorr), names(wcrcp45_uncorr), names(wcrcp26_uncorr)))

#set the extent for prediction
#e_asia = extent(50, 180, 15, 80) #not needed for Virginia can remove lines 129-163
#e_namerica = extent(-180, -70, 15, 80) #not needed for Virginia
#olivia change these values to the long/lat cutoff values for sampling: CHANGED on 2/3/22
e_Virginia = extent(-97.0000,-64.0000,25.0000,50.0000)

#creates new masked rasters for each based on extent
#wccurr_masked = raster_wrap_comb(e_asia, e_namerica, wccurr_uncorr)
#wclgm_masked = raster_wrap_comb(e_asia, e_namerica, wclgm_uncorr)
#wcrcp85_masked = raster_wrap_comb(e_asia, e_namerica, wcrcp85_uncorr)
#wcrcp60_masked = raster_wrap_comb(e_asia, e_namerica, wcrcp60_uncorr)
#wcrcp45_masked = raster_wrap_comb(e_asia, e_namerica, wcrcp45_uncorr)
#wcrcp26_masked = raster_wrap_comb(e_asia, e_namerica, wcrcp26_uncorr)

#changes longitudes to match new extent
#occ_masked = occ_distinct

#for (i in 1:length(occ_masked$Longitude)) {
  
  #case for points in e_asia
# if (occ_masked$Longitude[i] >= e_asia[1]) {
    
 #   occ_masked$Longitude[i] = occ_masked$Longitude[i] - e_asia[1]
#  }
  
  #case for points in e_namerica
 # else if (occ_masked$Longitude[i] <= e_namerica[2]) {
    
 #   occ_masked$Longitude[i] = occ_masked$Longitude[i] + 180 + (e_asia[2] - e_asia[1])
 # }
#}

#creates new extent of current range masked
#e_curr_range_masked = extent(min(occ_masked$Longitude) - 1, max(occ_masked$Longitude) + 1, min(occ_masked$Latitude) - 1, max(occ_masked$Latitude) + 1)

#exports masked occurence points, check file path
write.csv(occ_distinct, "C:/maxent/sangmodels/occ_masked_DRtest.csv")
#not sure what file I should replace this with- OM 2/3/22

```

runs projections 

```{r}

#ensures java is located for maxent
system.file("java", package="dismo")
file.copy("Maxent3.4.1/maxent.jar", "/Library/Frameworks/R.framework/Versions/3.0/Resources/library/dismo/java/maxent.jar")
jar <- paste(system.file(package="dismo"), "/java/maxent.jar", sep='')

#training and testing presence data, for Virginia replace "_masked" with "_distinct"
fold = kfold(occ_distinct, 5)
occtest = occ_distinct[fold == 1,]
occtrain = occ_distinct[fold != 1,]

#generates background samples, for Virginia change "_masked" to "_uncorr" and extent to VA variable
bg = randomPoints(wccurr_uncorr, n=1000, ext=e_Virginia)

#maxent with background samples
#parameters for maxent may need to be tweaked, for Virginia change "_masked" to "_uncorr"
mx = maxent(wccurr_uncorr, occtrain, a=bg, removeDuplicates=TRUE, args=c("-J") , path="C:/maxent/sangmodels/mxoutBR",)

plot(mx)

#predict and plot; for Virginia replace all "_masked" with "_uncorr"
modelcurr = predict(mx, wccurr_uncorr, ext=wccurr_uncorr, filename="C:/maxent/sangmodels/mxoutBR/BR_modelcurr.asc", overwrite=TRUE, progress='text')
modellgm = predict(mx, wclgm_uncorr, ext=wclgm_uncorr, filename="C:/maxent/sangmodels/mxoutBR/BR_modellgm.asc", overwrite=TRUE, progress='text')
modelrcp85 = predict(mx, wcrcp85_uncorr, ext=wcrcp85_uncorr, filename="C:/maxent/sangmodels/mxoutBR/BR_modelrcp85.asc", overwrite=TRUE, progress='text')
modelrcp60 = predict(mx, wcrcp60_uncorr, ext=wcrcp60_uncorr, filename="C:/maxent/sangmodels/mxoutBR/BR_modelrcp60.asc", overwrite=TRUE, progress='text')
modelrcp45 = predict(mx, wcrcp45_uncorr, ext=wcrcp45_uncorr, filename="C:/maxent/sangmodels/mxoutBR/BR_modelrcp45.asc", overwrite=TRUE, progress='text')
modelrcp26 = predict(mx, wcrcp26_uncorr, ext=wcrcp26_uncorr, filename="C:/maxent/sangmodels/mxoutBR/BR_modelrcp26.asc", overwrite=TRUE, progress='text')

#evaluates model, for Virginia change "_masked" to "_uncorr"
meval = evaluate(mx, p=occtest, a=bg, x=wccurr_uncorr)
```

```{r}
#gets threshold from model evaluation
#if using max sum of sensitivity and specificity, use spec_sens value
library(readr)

write_csv(threshold(meval), "C:/maxent/DIVI.csv")

# puts rasters in folder
writeRaster(modelcurr,'C:\\maxent\\BR_rasters/DIVImodelcurr2.tif', overwrite=TRUE)
writeRaster(modellgm,'C:\\maxent\\BR_rasters/DIVImodellgm2.tif', overwrite=TRUE)
writeRaster(modelrcp85,'C:\\maxent\\BR_rasters/DIVImodelrcp852.tif', overwrite=TRUE)
writeRaster(modelrcp60,'C:\\maxent\\BR_rasters/DIVImodelrcp602.tif', overwrite=TRUE)
writeRaster(modelrcp45,'C:\\maxent\\BR_rasters/DIVImodelrcp452.tif', overwrite=TRUE)
writeRaster(modelrcp26,'C:\\maxent\\BR_rasters/DIVImodelrcp262.tif', overwrite=TRUE)

```





